---
title: "02_stage1"
author: "Sujung Lee"
date: "2026-04-25"
output: html_document
---

# MCC STUDY — STAGE 1 (location-specific) ANALYSIS
Run this script LOCALLY on your hospitalisation data and return the RESULTS files (model coefficients + variances).

## How to use
1. Edit ONLY the CONFIGURATION block below:
 - file paths and output folder
 - your COUNTRY code
 - the CSV column separator
 - AGE_MAP  (map YOUR age-group labels to the harmonised codes)
2. Run the whole script:
3. Send back the four files written to `dir_out` (see OUTPUT below).

## REQUIRED HEALTH DATA FORMAT
- A column named exactly  date  (YYYY-MM-DD), one row per day.
- Every OTHER column is a daily hospitalisation count, named: <sex>_<age>_<cause>_<location>
where:
  <sex> = fem | mal (exactly these codes)
  <age> = 0to1 | 2to18 | 19to45 | 46to65 | 66to75 | 76to85 | more86 (or similar format)
  <cause> = one of CAUSES below
  <location> = city/area name (may contain spaces or "-")
  NOTE: sex, age and cause must NOT contain underscores; only <location> may.
    Examples:
        fem_19to45_cvd_mexico city
        mal_more86_resp_guadalajara

## Required output (<CC> is country code; United Kingdom = UK)
  <CC>_stage1_allsex_t2m.csv   all-sex, by cause
  <CC>_stage1_fem_t2m.csv       females, by cause
  <CC>_stage1_mal_t2m.csv       males, by cause
  <CC>_stage1_age_t2m.csv       age-specific (all-sex), by cause


#### Load libraries
```{r, echo = FALSE}
required <- c("dplyr", "tidyr", "lubridate", "stringr", "splines", "dlnm", "MESS")
missing  <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop("Missing packages. Please run:\n  install.packages(c(",
       paste(sprintf('"%s"', missing), collapse = ", "), "))")
}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(lubridate)
  library(stringr); library(splines); library(dlnm); library(MESS)
})
```

#### DLNM specification
```{r}
# Specification of the lag function
maxlag <- 21
lagfun <- "ns"
lagnk <- 3
lagknots <- logknots(maxlag, lagnk)

# Specification of the exposure function
varfun <- "ns" 
varper <- c(50)

predper <- c(seq(0, 1, 0.1), 2:98, seq(99, 100, 0.1))
axisper <- c(1, 25, 50, 75, 90)
mhprange <- c(25, 90)
minhosp <- 500

agebreaks <- c(20, 45, 65, 75, 85)
ageknots <- c(60)

causes <- c("all", "cvd", "resp", "inf", "ment", "uri")
```

## 1. Load data
```{r}
dir_data <- "path/to/your/file/directory/"

# Load climate data
dat_clim <- readRDS(paste0(dir_data, "MX_daily_list_t2m.rds"))

# Load health data
dat_health <- read.csv(paste0(dir_data, "mx_mcc_timeseries.csv"), sep = ";")
head(dat_health)
```

## Reformat health data to run 1st-stage analysis
Assumes the health data follow the MCC data protocol exactly:
  - one column named 'date' (YYYY-MM-DD)
  - every other column named  <sex>_<age>_<cause>_<location>
The age harmonisation below is FIXED by the protocol 
  — if you have different age groups, please contact Sujung Lee.
```{r}
dat_health_long <- dat_health %>%
  pivot_longer(
    cols = -date,
    names_to  = c("sex", "age", "cause", "location"),
    names_pattern = "^([^_]+)_([^_]+)_([^_]+)_(.+)$",
    values_to = "hosp"
  ) %>%
  mutate(age_regroup = case_when(
    age == "0to1" ~ "0018",
    age == "2to18" ~ "0018",
    age == "19to45" ~ "1945",
    age == "46to65" ~ "4675",
    age == "66to75" ~ "4675",
    age == "76to85" ~ "76up",
    age == "more86" ~ "76up",
    .default = age
  )) %>%
  mutate(
    date = as.Date(date),
        year = as.integer(format(date, "%Y")),
        month = as.integer(format(date, "%m")),
        dow = weekdays(date)
  ) %>%
  group_by(location, date, year, month, dow,
           sex, age_regroup, cause) %>%
  summarise(hosp = sum(hosp, na.rm=T), .groups = "drop") %>%
  mutate(col_name = paste0(sex, "_", age_regroup, "_", cause)) %>%
  dplyr::select(-sex, -age_regroup, -cause) %>%
  pivot_wider(
    names_from = col_name,
    values_from = hosp
  )

# Make sure names of location in health data and climate data are same
dat_health_long$location <- tolower(str_replace_all(dat_health_long$location, "[ -]", "_"))
dat_health_list <- dat_health_long %>% split(.$location)
rm(dat_health_long)
```

## Merge health and climate data
```{r}
dlist <- list()
for (i in seq_along(names(dat_health_list))) {
      
  # Extract city name
  city_name <- names(dat_health_list)[i]
  
  # Extract climate and health data for selected city
  dat_health_city <- dat_health_list[[city_name]]
  start_date <- as.Date(min(dat_health_city$date))
  end_date <- as.Date(max(dat_health_city$date))
  dat_clim_city <- dat_clim[[city_name]] %>%
    mutate(date = as.Date(date)) %>%
    filter(date >= start_date & date <= end_date)
  
  # Merge health and climate data
  dat_city <- left_join(dat_clim_city, dat_health_city, by = c("location", "date"))
  dat_city <- dat_city %>%
    mutate(across(7:ncol(.), ~replace_na(.x, 0)))
  
  dlist[[city_name]] <- dat_city
}

# Check NA
na_check <- lapply(dlist, function(df) {
  na_counts <- colSums(is.na(df))
  na_counts[na_counts > 0]
})

print(na_check)

rm(dat_health_list, dat_clim)
```

## All-age analysis
#### 1st stage: Fit location-cause models by aggregating all age groups
```{r}
stage1_all <- list()

for (cx in causes) {
    
  # Create empty list to store results for selected cause
  stage1_all[[cx]] <- list()
    
  for (i in seq_along(names(dlist))) {
      
    # Extract city name
    city_name <- names(dlist)[i]
    
    # Extract main data for selected months
    city_df <- dlist[[city_name]] %>%
      mutate(
        date = as.Date(date),
        year = year(date),
        month = month(date),
        dow = wday(date, label = TRUE))
  
    varknots <- quantile(city_df$t2m, varper / 100, na.rm = TRUE)
    boundknt <- range(city_df$t2m, na.rm = TRUE)
    
    argvar <- list(fun = varfun, knots = varknots, Bound = boundknt)
    cb <- crossbasis(city_df$t2m,
                     lag = maxlag,
                     argvar = argvar,
                     arglag = list(fun = lagfun, knots = lagknots))
    
    # Extract column names for all_sex, females, and males
    yvars_all <- grep(sprintf("^(fem|mal)_[0-9a-z]+_%s$", cx), names(city_df), value = TRUE)
    yvars_fem <- grep(sprintf("^(fem)_[0-9a-z]+_%s$", cx), names(city_df), value = TRUE)
    yvars_mal <- grep(sprintf("^(mal)_[0-9a-z]+_%s$", cx), names(city_df), value = TRUE)
    
    if (length(yvars_all) == 0) {
      message("No outcome variables found for city: ", city_name, " | Outtype: ", cx)
      next
    }
      
    city_df$y_all <- rowSums(city_df[, yvars_all, drop = FALSE], na.rm = TRUE)
    city_df$y_fem <- rowSums(city_df[, yvars_fem, drop = FALSE], na.rm = TRUE)
    city_df$y_mal <- rowSums(city_df[, yvars_mal, drop = FALSE], na.rm = TRUE)
      
    nred <- attr(cb, "df")[1]
    nvcov <- sum(1:nred)
    
    cnames_coef <- paste0("b", seq(nred))
    cnames_vcov <- sprintf("v%i%i", 
      row(matrix(0, nred, nred))[row(matrix(0, nred, nred)) >= col(matrix(0, nred, nred))],
      col(matrix(0, nred, nred))[row(matrix(0, nred, nred)) >= col(matrix(0, nred, nred))])
    
    cat("Exposure: t2m | Outtype:", cx, "| City:", city_name, 
        "| Total cases:", sum(city_df$y_all, na.rm = TRUE), 
        "| Female cases:", sum(city_df$y_fem, na.rm = TRUE),
        "| Male cases:", sum(city_df$y_mal, na.rm = TRUE), "\n")
        
    n_years <- length(unique(city_df$year))
    
    yvars_list <- c(all_sex = "y_all", fem = "y_fem", mal = "y_mal")
    
    for(sx in names(yvars_list)) {
      
      yv <- yvars_list[[sx]]
      fmla <- as.formula(paste0(yv, " ~ cb + dow + ns(date, df = 7 * n_years)"))

      res <- tryCatch(
        withCallingHandlers(
          glm(fmla, city_df, family = quasipoisson),
          warning = function(w) {
            cat("GLM warning in ", city_name, ":", conditionMessage(w), "\n")
            invokeRestart("muffleWarning")
          }
        ),
        error   = function(e) { 
          message("GLM error in ", city_name, ": ", e$message) 
          NULL 
          }
        )
      
      if (is.null(res) || !res$converged) {
        cat("Not converged:", city_name, "| Cause:", cx, "| Sex:", sx, "\n")
        na_coef <- setNames(rep(NA, nred), cnames_coef)
        na_vcov <- setNames(rep(NA, nvcov), cnames_vcov)
        stage1_all[[cx]][[sx]][[city_name]] <- data.frame(
          location = city_name,
          exposure = "t2m",
          t(na_coef), 
          t(na_vcov),
          tothosp = sum(city_df[[yv]], na.rm = T), 
          conv = FALSE,
          sex = sx,
          cause = cx
          )
          next
        }
        
      redall <- crossreduce(cb, res, cen = median(city_df$t2m, na.rm = TRUE))
      trimat <- row(vcov(redall)) >= col(vcov(redall))
      vvec <- vcov(redall)[trimat]
      names(vvec) <- sprintf("v%i%i", 
                             row(vcov(redall))[trimat], 
                             col(vcov(redall))[trimat])
      
      stage1_all[[cx]][[sx]][[city_name]] <- data.frame(
        location = city_name,
        exposure = "t2m",
        t(coef(redall)), 
        t(vvec),
        tothosp = sum(city_df[[yv]], na.rm = T), 
        conv = res$converged,
        sex = sx,
        cause = cx
      )
    }
  }
}

allsex_df <- do.call(rbind, lapply(causes, function(cx) do.call(rbind, stage1_all[[cx]][["all_sex"]])))
fem_df <- do.call(rbind, lapply(causes, function(cx) do.call(rbind, stage1_all[[cx]][["fem"]])))
mal_df <- do.call(rbind, lapply(causes, function(cx) do.call(rbind, stage1_all[[cx]][["mal"]])))

# Save results
dir_output <- "/path/to/your/output/directory"

write.csv(allsex_df, 
          file = paste0(dir, "MX_project1_stage1_allsex_t2m.csv"), 
          row.names = F)

write.csv(fem_df, 
          file = paste0(dir, "MX_project1_stage1_fem_t2m.csv"), 
          row.names = F)

write.csv(mal_df, 
          file = paste0(dir, "MX_project1_stage1_mal_sex_t2m.csv"), 
          row.names = F)
```

## Age-specific analysis
#### 1st stage: location-age-cause estimates
```{r}
stage1_age <- list()
 
for (outtype in causes) {
  
  stage1_age[[outtype]] <- list()
  
  for (i in seq_along(names(dlist))) {
    
    city_name <- names(dlist)[i]
    
    city_df <- dlist[[city_name]] %>%
      mutate(
        date = as.Date(date),
        year = year(date),
        month = month(date),
        dow = wday(date, label = TRUE)
        )
    
    varknots <- quantile(city_df$t2m, varper / 100, na.rm=TRUE)
    boundknt <- range(city_df$t2m, na.rm=TRUE)
      
    argvar <- list(fun = varfun, knots = varknots, Bound = boundknt)
    cb <- crossbasis(city_df$t2m,
                     lag = maxlag,
                     argvar = argvar,
                     arglag = list(fun = lagfun, knots = lagknots))
      
    yvars <- grep(sprintf("^(fem|mal)_[0-9a-z]+_%s$", outtype), 
                  names(city_df), value = TRUE)
    
    if (length(yvars) == 0) {
      message("No outcome variables found for city: ", 
              city_name, " | outtype: ", outtype)
      next
      }
    
    # combine fem + mal within each age group
    sxage   <- gsub("^(fem|mal)_([0-9a-z]+)_.*", "\\2", yvars)  # age per column
    age_levels <- unique(sxage)
    age_levels <- age_levels[order(as.numeric(substr(age_levels, 1, 2)))]  # ascending by amin

    combvars <- character(0)
    for (ag in age_levels) {
      cols_ag <- yvars[sxage == ag]                       # fem + mal for this age
      newcol  <- sprintf("allsex_%s_%s", ag, outtype)         # e.g. all_0014_all
      city_df[[newcol]] <- rowSums(city_df[, cols_ag, drop = FALSE], na.rm = TRUE)
      combvars <- c(combvars, newcol)
    }
    
    # Extract age groups from the column name
    ageparts <- gsub("^allsex_([0-9a-z]+)_.*", "\\1", combvars)
    is_openended <- grepl("[a-z]", substr(ageparts, 3, 4))
    amin <- substr(ageparts, 1, 2)
    amax <- ifelse(is_openended, "99", substr(ageparts, 3, 4))
      
    ytot <- apply(city_df[, combvars, drop = FALSE], 2, sum, na.rm = TRUE)
    cumgroups <- cumsumbinning(ytot, minhosp, cutwhenpassed = TRUE)
      
    # Check the number of cases in the last group
    last_group <- max(cumgroups)
    last_total <- sum(ytot[cumgroups == last_group])
 
    # If the number of cases in the last age group is < minhosp, merge it with the preceding age group
    if (last_total < minhosp && last_group > 1) {
      cumgroups[cumgroups == last_group] <- last_group - 1
      cat("Merged last age group in", city_name, 
          "| cases:", last_total, "-> combined with group", last_group - 1, "\n")
      }
      
    nred <- attr(cb, "df")[1]
    nvcov <- sum(1:nred)
      
    cnames_coef <- paste0("b", seq(nred))
    cnames_vcov <- sprintf("v%i%i", 
      row(matrix(0, nred, nred))[row(matrix(0, nred, nred)) >= col(matrix(0, nred, nred))],
      col(matrix(0, nred, nred))[row(matrix(0, nred, nred)) >= col(matrix(0, nred, nred))])
      
    ageres <- list()
    
    for (a in unique(cumgroups)) {
      avars <- combvars[cumgroups == a]
      city_df$y <- rowSums(city_df[, avars, drop = F])
      aamin <- amin[cumgroups == a][1]
      aamax <- tail(amax[cumgroups == a], 1)
      meanage <- (as.numeric(aamin) + as.numeric(aamax)) / 2
        
      cat("Exposure: t2m | Outtype:", outtype, "| City:", city_name, 
          "| Age:", aamin, "-", aamax, "\n")
        
      n_years <- length(unique(city_df$year))

      res <- tryCatch(
        withCallingHandlers(
        glm(y ~ cb + dow + ns(date, df = 7 * n_years),
            city_df, family = quasipoisson),
        warning = function(w) {
          cat("GLM warning in ", city_name, ":", conditionMessage(w), "\n")
          invokeRestart("muffleWarning")
          }
        ),
        error   = function(e) { 
          message("GLM error in ", city_name, ": ", e$message)
          NULL 
          }
        )
        
      if (is.null(res) || !res$converged) {
        cat("Not converged:", city_name, "| Age:", aamin, "-", aamax, 
            "| Cause:", outtype, "\n")
        na_coef <- setNames(rep(NA, nred), cnames_coef)
        na_vcov <- setNames(rep(NA, nvcov), cnames_vcov)
        ageres[[a]] <- data.frame(
          t(na_coef), 
          t(na_vcov),
          tothosp = sum(city_df$y, na.rm = T), 
          conv = F,
          ageval = meanage, 
          agegr = paste0(aamin, aamax), 
          cause = outtype, 
          n_agegroups = length(unique(cumgroups))
          )
        next
      }
        
      redall <- crossreduce(cb, res, cen = median(city_df$t2m, na.rm = TRUE))
      trimat <- row(vcov(redall)) >= col(vcov(redall))
      vvec <- vcov(redall)[trimat]
      names(vvec) <- sprintf("v%i%i", row(vcov(redall))[trimat],
                             col(vcov(redall))[trimat])
        
      ageres[[a]] <- data.frame(
        t(coef(redall)), 
        t(vvec),
        tothosp = sum(city_df$y, na.rm = T), 
        conv = res$converged,
        ageval = meanage, 
        agegr = paste0(aamin, aamax), 
        cause = outtype,
        n_agegroups = length(unique(cumgroups))
        )
    }
      
    stage1_age[[outtype]][[city_name]] <- cbind(
      location = city_name, do.call(rbind, ageres))
   }
}
  
stage1_age_combined <- lapply(stage1_age, function(x) do.call(rbind, x))
stage1_age_combined_all <- do.call(rbind, stage1_age_combined)
rownames(stage1_age_combined_all) <- NULL

# Save results
dir_output <- "/path/to/your/output/directory/"
write.csv(stage1_age_combined_all, 
          file = paste0(dir, "MX_project1_stage1_age_t2m.csv"), 
          row.names = F)
```
